DySECT -- truly space efficient dynamic hashing

### 2017-04-18 paper/paper.tex
Work - After Sebastians Improvements

% Author macros::begin %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{DySECT -- truly space efficient dynamic hashing} % -- a novel approach to growing!}
\titlerunning{A Sample LIPIcs Article} %optional, in case that the title is too long; the running title should fit into the top page column

@@

\begin{abstract}
Research has concentrated on optimizing the speed and space efficiency
of hash tables for a long time.  Modern open addressing techniques can
fill a given table to more than 95\% before the performance
deteriorates significantly.  But once the threshold is met, copying
and migrating the whole table still seems to be the most efficient
answer.  Growing by a constant factor $\gamma$ often leads to bad
space efficiency because the newly grown table is only filled less
than a factor of $\gamma^{-1}$.

The space efficiency is even worse during the migration, where
oftentimes both the new and the old table coexist at the same time
causing a massive space overhead.  To solve both of these issues, we
propose the a new hash table architecture that supports dynamic
growing while keeping a strict size constraint $\alpha \cdot n$
throughout its lifetime.  Even under these conditions, we achieve
guaranteed constant lookup times, and amortized insertion times,
independent of $n$.

We achieve this goal by splitting the hash table into subtables that
grow independent from each other.  The cuckoo load balancing mechanism
can move elements from one table to the other.  This ensures that
growing one subtable increases insertion opportunities even for
elements that are not hashed into this one table.
 \end{abstract}

@@

\section{Introduction}
Hash tables are some of the most frequently used data-structures. They
belong into every programmers basic toolbox.  As such having
interchangeable implementations which perform well under different
conditions is important and allows programmers to easily adapt known
solutions to new circumstances.

One aspect which a lot of research focuses on is space efficiency.
Modern space efficient hash tables work well up to loads of 95\% and
more. But for this to matter, programmers have to know similarly tight
bounds to the absolute number of elements inserted into the table.
This is typically not the case, one of the most ubiquitous use cases
for hash tables is to unify data elements by their key.  To guarantee
good performance, we have to overestimate the initial capacity
whenever the exact number of unique keys is not known a priori.
Dynamic data-structures are necessary to alleviate this problem.  In
circumstances where space efficiency is important, we need space
efficiently growing hash tables.

Many libraries -- even ones that implement space efficient hash tables
-- already offer dynamic growing.  The problem with these
implementations is that they either lose their space efficiency or
their overall performance, once the table grows above its original
capacity.  Usually growing is implemented by either creating
additional hash tables -- decreasing the performance especially for
look ups, or by migrating all elements to a new table -- losing the
space efficiency by multiplying the original size.

To avoid both of these pitfalls we propose a variant of bucket cuckoo
hashing (TODO REFERENCE).  Bucket cuckoo hashing is a technique where
each element can be stored in one of several associated constant sized
buckets.  When all of them are full we move elements, to create new
space.  To solve the problem of efficient migration, we split the
table into multiple subtables each of which can grow independently
from all others.  Because the buckets associated with one element are
spread over the different subtables, growing one subtable alleviates
pressure from all others by allowing moves from a dense table to the
grown table.

@@

\section{Motivation}
\label{sec:motivation}
Space efficiency is a major factor for the development of new hashing
techniques.  The goal has long been to make hash tables work better in
high load scenarios.  Modern hashing techniques like Hopscotch, Robin
Hood, and Cuckoo Hashing can easily fill hash tables to more than 90\%
and up to 98\% while still achieving constant average running times on
most of their operations.  But since these techniques depend on
statically sized tables this only makes sense, if the number of unique
elements can be estimated with similarly tight bounds.

To visualize this assume the following scenario.  During the operation
of a word count benchmark, we know an upper bound $c$ to the number of
unique words. Therefore, we construct a table large enough to hold
these elements ($0.9^{-1}\cdot c$ cells).  It is entirely possible
that the specific word count instance only contains $0.7\cdot c$
unique words (or less).  The resulting table is therefore only filled
63\%.  It is clear that better static sized tables cannot achieve fill
ratios greater than 70\%.

Depending on the underlying problem it is impossible to tightly
estimate the final number of elements. Since dynamic growing is independent from the problem's domain, it can be a
solution even in these instances.  To achieve space efficiency in
scenarios where the final size is not known the hash table has to grow
closely with the actual number of elements.  This is not efficient
with any of the current techniques used for hashing and migration.

@@

Over the last one and a half decades, the field has regained
attention, both from theoretical and the practical point of view.  The
initial innovation that sparked this attention was the power of two
choices~\cite{ThePowerOfTwoChoicesInRandomizedLoadBalancing}.  This
concept led to the development of cuckoo hashing~\cite{CuckooHashing}
which reopened research into more space efficient hash tables.  They
proofed theoretically interesting, because probabilistic bounds for
the maximum fill degree are often highly non-trivial TODO SOME BOUNDS
PAPERS.

Cuckoo hashing also spawned many variants, that further improve its
performance.  The most notable one is probably bucket cuckoo hashing
($d$-ary cuckoo hashing), a variant that groups cells together in
buckets thus improving cache usage, and increasing displacement
opportunities TODO BUNCH OF PAPERS.  This further improves the maximum
fill degree. Nowadays, bucket cuckoo hashing is by far the most common
variant. There are even further variations on bucket cuckoo hashing,
like the generalization to continuous buckets TODO PAPER.

Further adaptations to cuckoo hashing include: a de-amortization
technique that allows a provable worst case analysis of
insertions~\cite{UsingAQueueToDeAmortizeCuckooHashingInHardware,
  DeAmortizedCuckooHashingProvableWorstCasePerformanceAndExperimentalResults};
and multiple concurrent implementations either powered by bucket
locking, transactional
memory~\cite{AlgorithmicImprovementsForFastConcurrentCuckooHashing},
or fully lock-less~\cite{LockFreeCuckooHashing}.

Some non-cuckoo space efficient hash tables continue to use linear
probing variants.  Robin Hood hashing is a technique that was
originally introduced in 1985~\cite{RobinHoodHashing}. It has regained some popularity in
recent years -- mainly for its interesting theoretical
properties and the possibility to reduce the inherent variance of
linear probing.  Hopscotch~\cite{HopscotchHashing} hashing is a
technique, that originated in the research of concurrent hash tables, but
it can also lead to some densely filled hash tables.  For this reason
it is often used as a technique to achieve space efficient hash
tables.  This is only possible in a small number of circumstances, as
the stored bitmaps used in hopscotch hashing create a significant
memory overhead.

All these publications show that there is a clear interest in the
development of space efficient hash tables.  Dynamic hash tables on the
other hand seem to be considered a solved problem.  There is an
important paper by Dietzfelbinger at
al.~\cite{DynamicPerfectHashingUpperAndLowerBounds} that takes on the
problem of dynamic hash tables from a theoretical perspective.  The
paper predates cuckoo hashing, and much of the attention for space
efficient hashing. Thus all memory bounds presented are given without
tight constant factors.  This lack of development is where we pick up
and offer fast solutions that support dynamic growing with tight space
bounds.

\section{Preliminaries}
A hash table is a data-structure, which stores key-value-pairs
($\langle key, data \rangle$) and offers the following functions:
\verb~insert~ -- stores a given key-value pair; \verb~find~ -- given a
key returns if said key was stored, and $\bot$ otherwise; and
\verb~erase~ -- deletes a previously inserted element (if present).

@@

Throughout this paper, we use $n$ to denote the number of elements and
$m$ the number of cells ($m > n$) in any given hash table.  We define
the load factor as $load = n/m$, and we use $\varepsilon = 1-load$ for
the fraction of available cells.  Tables can usually only work
efficiently up to a certain load factor.  Above that, operations get
slower and/or have a possibility to fail.  When implementing a hash
table one has to decide between storing elements directly in the hash
table -- \emph{Closed Hashing} -- or storing pointers to elements --
\emph{Open Hashing}. This has an immediate impact on the necessary
memory.

\begin{align*}
  mem_{closed}(n,m) &\geq m\cdot size(element)\\
  mem_{open}  (n,m) &\geq m\cdot size(pointer) + n\cdot size(element)\\
\end{align*}

For large elements ($\gg size(pointer)$), one can use a non-space
efficient hash table with open hashing, to reduce the relevant memory
factor.  Therefore, we restrict ourselves to the common and more interesting case of
elements whose size is comparable to that of a pointer.  For our
experiments we use 128bit elements (64bit keys and 64bit values).  In
these instances open hashing introduces a significant memory
overhead (at least $\times1.5$), therefore, we only consider closed
hash tables for the purpose of this paper. For closed hash tables, the
memory efficiency is directly dependent on the element density.  To
reach high fill degrees with closed hashing tables, we have to employ
\emph{open addressing} techniques which means, that elements are not
stored in predetermined cells, but can be stored in one of several
options (e.g. linear probing, or cuckoo hashing).

\subsection{$\alpha$-Space Efficient Hash Tables}
\subparagraph*{Static}
\label{sec:pre_staticspace}
We call a hashing technique $\alpha$-space efficient when it can work
effectively using at most $\alpha \cdot n_{\textit{curr}} \cdot \textit{size}(\textit{element}) + O(1)$ memory. In
this case we define working efficiently as having average insertion
times close to $O(\frac{1}{\varepsilon})$. This is a natural bound
since it is the expected number of fully random probes needed to hit
an empty cell.

@@

In many closed hashing techniques (e.g.~linear probing, cuckoo
hashing) cells are the same size as elements. Therefore, being
$\alpha$ space efficient is the same as operating with a load factor
of $\alpha^{-1} = n/m$.
%(because $\alpha\cdot n \cdot size(element) = m\cdot size(element)$).
Because of this, we will mostly talk about the load factor of a table
instead of its memory usage.

@@

\paragraph*{Dynamic}
The definition of a space efficient hashing technique given aboce is
specifically targeted for statically sized hash tables.  We call a
hash table implementation dynamically $\alpha$-space efficient if an
instantiated table can grow arbitrarily large over its original
capacity, while remaining smaller than $\alpha\cdot n_{\max}\cdot
\textit{size}(\textit{element}) + O(1)$ at all times.
%\[mem_{curr} \leq \alpha\cdot max(n_{curr})\cdot size(element)\]

One problem for many implementations of space efficient hash tables is
the migration.  During a normal full table migration, both the
original table, and the new table are allocated.  This takes
$\textit{m}_{\textit{new}} + \textit{m}_{\textit{old}}$
cells. Therefore, no normal full table migration can be more then
2-space efficient.  The only option to perform a full table migration
with less memory, is to increase the memory in place (see
Section~\ref{sec:dys_inplace} and
Section~\ref{sec:exp_competitor}).

Similar to static $\alpha$-space efficiency, we will mostly talk about
the \emph{minimum load factor}, instead of $\alpha$.  We should note,
that this definition of dynamic $\alpha$-space efficient hash tables
does not enforce size reduction due to the deletion of elements, but
it enforces that cells previously used by deleted elements must be reused.

\subsection{Cuckoo Hashing}
\label{sec:pre_cuckoo}
Cuckoo hashing is a technique to resolve hash conflicts in a closed
hashing hash table. Its main draw is that it guarantees constant
lookup times even in densely filled tables.  The distinguishing
technique of cuckoo hashing is, that $H$ hash functions ($h_1, ... ,
h_H$) are used to compute $H$ independent positions. Each element is
stored in one of its positions.  Even if all positions are occupied
one can often move elements to create space for the current
element. We call this process displacing elements.

Bucket cuckoo hashing is a variant, where the cells of the hash table
are grouped into buckets of size $B$ ($m/B$ buckets).  Each element
appointed to one bucket can be stored in any of the bucket's cells .
Using buckets one can drastically increase the number of displacement
opportunities.

\emph{Find} and \emph{remove} operations have a guaranteed constant
running time. Independent from the tables density, there are $H$
buckets -- $H\cdot B$ cells -- that have to be searched to find an
element.

During an \emph{insert} the element is hashed to $H$ buckets.  If at least
one of those buckets has an empty cell, we store the element in the
bucket with the most space.  When all buckets are full, then we
have to move elements around the table, such that a new space becomes
available.

\label{sec:cuckoo_graph}
To visualize the problem of displacing elements, one can think of the
directed graph implicitly defined by the hash table.  Each bucket
defines one node and each element defines pairwise connections between
the bucket it is stored in to its $H-1$ alternate buckets.  To insert
an element into the hash table we have to find a path from one of its
associated buckets, to a bucket, that has space remaining.  Then we
move elements along this path, to open a space on the initial bucket.
The two common techniques to find such paths are \emph{random walks} and
\emph{breadth first searches}.

\section{DySECT Space Efficient Growing, by Splitting the Table}
Our goal is to build an efficiently growing data structure, that
remains space efficient at all times.  A commonly used growing
technique is to double the size by migrating all elements
into a larger new table.  Of course this is
not memory efficient.  A natural thought is to double
only part of the data structure.  This idea together with the
balancing mechanism of cuckoo hashing will achieve the
described functionality.

\subsection{Description}
We use $T$ subtables (shown in TODO FIGREFRENCE) to allow
differentiated growing between table parts.  The subtables consist of
buckets that store elements.

Each element has $H$ associated
buckets similar to cuckoo hashing, these can be in the same, or
in different subtables.  To find a bucket associated with an element $e$,
we compute $e$'s hash value using the appropriate hash function
$h_i(e)$, from a hashed key, we have to compute the subtable and the bucket
within that subtable.

To make this efficient we use powers of two for the number of
subtables ($T = 2^t$), as well as the individual subtable sizes ($s =
2^x*B$).  Since the number of subtables is constant we can use the
first $t$ bits from the hashed key, to find the appropriate subtable.
From the remaining bits we compute the offset into this subtable using
a bitmask ($h_i(e)\,\texttt{\&}\,(s-1) = h_i(e) \mod s$).

@@

\subsection{Growing}
We grow one subtable,
when the table contains enough elements, that the memory constraint
can be kept during the migration.  We migrate subtables in order
from the first to the last, therefore, no subtable can be more than
twice as large as any other.

Assume that all subtables have $s=m/T$ cells. When $\alpha\cdot n >
m+2s$ we can grow the first subtable while keeping the size
constraint.  Doubling the size of a subtable increases the global
number of cells from $m_{old} = T\cdot s$ to $m_{new} = m_{old}+s =
(T+1)\cdot s$ (grow factor $\frac{T+1}{T}$).  Note that all subsequent
grows migrate one of the smaller tables, until all tables have the
same size again.  Therefore, each grow until then increase the
capacity by the same absolute amount (smaller grow factor).

The cost of growing the subtable is amortized by all insertions that
happened since the last subtable migration.  There are at least
$\alpha^{-1} \cdot s = \Omega(s)$ insertions between two migrations.
One migration takes $\Theta(s)$ time.  Apart from being amortized,
the migration is efficient in practice because it accesses cells in a
linear fashion making it really cache efficient.  Even in the target
table cells are accessed linearly.  Because of the way we assign
elements to buckets, one original bucket is split into two target
buckets.  Therefore, no displacements are necessary, since no bucket
can be overfilled.

In the implicit graph model of the cuckoo table
(Section~\ref{sec:cuckoo_graph}), growing a subtable is equivalent to
splitting each node that represents a bucket within that table. The
edges (elements) in the implicitly defined subgraph are not doubled,
therefore, the resulting subgraph becomes more sparse, making it
easier to insert elements.

\subsection{Shrinking}
In many use cases, shrinking is not necessary.  It worsens performance
by taking time for the migration and making the remaining table more
dense.  Furthermore, it is unclear that the freed memory is actually
necessary for other parts of the application.  If that memory is
necessary, and shrinking is required, for example, when removed
elements are reinserted into another data structure and global memory
bounds are necessary (other data structures grow while the hash table
shrinks).

Shrinking can work similarly to growing. We replace a subtable with a
smaller one by migrating elements from one to the other.  During this
migration we join elements from two buckets into one. Therefore, it is
possible for a bucket to overfill.  We reinsert these elements at the
end of the migration.  Obviously, this can only apply to at most half
the migrated elements.

When triggering the size reduction, one has to make sure that the
migration is amortized. Therefore, a grow operation cannot immediately
follow a shrink operation.  When shrinking is enabled we propose to
shrink one subtable when $\alpha\cdot n < m-s'$ elements ($s'$ size of a
large table, $m_{new} = m_{old} -s'/2$).

@@

\subsection{Implementation Details}
For our experiments in Section~\ref{sec:exp} we use three hash
functions ($H=3$) and a bucket size of ($B=8$). These values have
consistently outperformed other options.  We chose to use $T=256$
subtables for all our tests.  The number of subtables only plays a
minor role for the running time.  To find displacement opportunities
we use breadth first search.  In our tests it performed better than
random walks, since it better uses the read cache lines from one
bucket.

\subsubsection*{Reducing Number Computed Hash Functions}
Evaluating hash functions is expensive, therefore, reducing the number
of hash functions computed per operation increases the efficiency of
the table.  The hash function we use computes 64bit hash values
(i.e. xxHash TODO).  We split the 64bit value into two 32bit hash
values. All common bucket hash table sizes can be addressed using 32
bits. They can address up to $2^{32}$ buckets which achieves a
capacity for $2^{35} \approx 34$ billion elements (bucket size 8),
at that point a hash table consumes 512GiB memory.

When $H > 2$ we can use \emph{double hashing} to further reduce the number
of computed hash functions. Double hashing creates an arbitrary amount
of hash values using only two original hash functions $h'$ and
$h''$.  The additional values are linear combinations computed
from the original two values, $h_i(key) = h'(key) + i\cdot
h''(key)$.

>>>>

% Author macros::begin %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{DySECT -- truly space efficient dynamic hashing} % -- a novel approach to growing!}
\titlerunning{DySECT -- truly space efficient dynamic hashing} %optional, in case that the title is too long; the running title should fit into the top page column

@@

\begin{abstract}
Research has concentrated on optimizing the speed and space efficiency
of hash tables for a long time.  Hash tables using modern open
addressing techniques can be filled to more than 95\% before the
performance deteriorates significantly.  Once this threshold is met,
copying and migrating the whole table still seems to be the most
efficient answer.  Growing by a constant factor $\gamma$ often leads
to bad space efficiency because the filling degree of the grown table
is at most $\gamma^{-1}$. During the migration, the space efficiency
is even worse, since oftentimes both the new and the old table coexist
at the same time, causing a massive space overhead.

We propose a new hash table
architecture that supports dynamic growing while keeping a strict size
constraint $\alpha \cdot n$ throughout its lifetime.  Even under these
conditions, we achieve guaranteed constant lookup times, and amortized
constant insertion times.

Our architecture splits the overall hash table into subtables that
grow independent from each other.  The cuckoo load balancing mechanism
can move elements between subtables ensuring that growing one
subtable increases the overall insertion opportunities.
 \end{abstract}

@@

\section{Introduction}
Hash tables are some of the most frequently used data-structures. Having
interchangeable implementations which perform well under different
conditions is important and allows programmers to easily adapt known
solutions to new circumstances.

One aspect a lot of research focuses on is space efficiency TODO
REFERENCES.  Modern space efficient hash tables work well up to loads
of 95\% and more.  But for this to matter, the table has to be
initialized with the correct final capacity, therefore, forcing
programmers to know tight bounds to the maximum number of inserted
elements.  This is typically not the case.  One of the most ubiquitous
use cases for hash tables is to unify data elements by their key.
Whenever the exact number of unique keys is not known a priori we have
to overestimate the initial capacity to guarantee good performance.
Dynamic space efficient data-structures are necessary to alleviate
this problem independent of the circumstances.

To visualize this assume the following scenario.  During the operation
of a word count benchmark, we know an upper bound $c$ to the number of
unique words. Therefore, we construct a table large enough to hold
these elements ($0.9^{-1}\cdot c$ cells).  It is entirely possible
that the specific word count instance only contains $0.7\cdot c$
unique words.  The resulting table is therefore only filled to 63\%.
It is clear that better static sized tables cannot achieve fill ratios
greater than 70\%.  A factor of 0.7 seems realistic considering
that the benchmark from Section~\ref{sec:exp_wordcount} for example
contains 240\,000\,000 words only 20\,000\,000 of which are unique.

Depending on the underlying problem it is impossible to tightly
estimate the final number of elements. Since growing the table
dynamically is independent from the problem domain, it can be a
solution even in these instances.  To achieve space efficiency in
scenarios where the final size is not known the hash table has to grow
closely with the actual number of elements.  This cannot be achieved
efficiently Swith any of the current techniques used for hashing and
migration.

Many libraries -- even ones that implement space efficient hash tables
-- already offer growing.  The problem with these implementations is
that they either lose their space efficiency or their performance
degrades once the table grows above its original capacity.  Usually
growing is implemented by either creating additional hash tables --
decreasing the performance especially for look ups, or by migrating
all elements to a new table -- losing the space efficiency by
multiplying the original size.

To avoid both of these pitfalls we propose a variant of bucket cuckoo
hashing (TODO REFERENCE).  Bucket cuckoo hashing is a technique where
each element can be stored in one of several associated constant sized
buckets.  When all of them are full we move an element into another of
its buckets to create new space.  To solve the problem of efficient
migration, we split the table into multiple subtables each of which
can grow independently from all others.  Because the buckets
associated with one element are spread over the different subtables,
growing one subtable alleviates pressure from all others by allowing
moves from a dense subtable to the grown subtable.

@@

%% \section{Motivation}
%% \label{sec:motivation}
%% Space efficiency is a major factor for the development of new hashing
%% techniques.  The goal has long been to make hash tables work better in
%% high load scenarios.  Modern hashing techniques like Hopscotch, Robin
%% Hood, and Cuckoo Hashing can easily fill hash tables to more than 90\%
%% and up to 98\% while still achieving constant average running times on
%% most of their operations.  But since these techniques depend on
%% statically sized tables this only makes sense, if the number of unique
%% elements can be estimated with similarly tight bounds.

%% To visualize this assume the following scenario.  During the operation
%% of a word count benchmark, we know an upper bound $c$ to the number of
%% unique words. Therefore, we construct a table large enough to hold
%% these elements ($0.9^{-1}\cdot c$ cells).  It is entirely possible
%% that the specific word count instance only contains $0.7\cdot c$
%% unique words (or less).  The resulting table is therefore only filled
%% 63\%.  It is clear that better static sized tables cannot achieve fill
%% ratios greater than 70\%.

%% Depending on the underlying problem it is impossible to tightly
%% estimate the final number of elements. Since dynamic growing is
%% independent from the problem's domain, it can be a solution even in
%% these instances.  To achieve space efficiency in scenarios where the
%% final size is not known the hash table has to grow closely with the
%% actual number of elements.  This is not efficient with any of the
%% current techniques used for hashing and migration.

@@

Over the last one and a half decades, the field has regained
attention, both from theoretical and the practical point of view.  The
initial innovation that sparked this attention was the idea, that
storing an element in the less filled of two ``random'' chains leads to
incredibly well balanced loads.  This concept is called the power of
two choices~\cite{ThePowerOfTwoChoicesInRandomizedLoadBalancing}.

It led to the development of cuckoo hashing~\cite{CuckooHashing}.
Cuckoo hashing extends the power of two choices by allowing to move
elements within the table when all chosen positions are already
occupied (see Section~\ref{sec:pre_cuckoo} for a more elaborated
explanation).  Cuckoo hashing revitalized research into space
efficient hash tables.  Probabilistic bounds for the maximum fill
degree and expected displacement distances are often highly
non-trivial TODO SOME BOUNDS PAPERS.

Cuckoo hashing also spawned many variants that further improve its
performance.  The most notable one being bucket cuckoo hashing
($d$-ary cuckoo hashing) a variant that groups cells together in
buckets thus improving cache usage, increasing displacement
opportunities and maximum fill degree TODO BUNCH OF PAPERS.  There are
even further variations on bucket cuckoo hashing, like the
generalization to continuous buckets TODO PAPER.

Further adaptations of cuckoo hashing include:multiple concurrent implementations either powered by bucket
locking, transactional
memory~\cite{AlgorithmicImprovementsForFastConcurrentCuckooHashing},
or fully lock-less~\cite{LockFreeCuckooHashing}; and a de-amortization
technique that allows a provable worst case analysis of
insertions~\cite{UsingAQueueToDeAmortizeCuckooHashingInHardware,
  DeAmortizedCuckooHashingProvableWorstCasePerformanceAndExperimentalResults}.

Some non-cuckoo space efficient hash tables continue to use linear
probing variants.  \emph{Robin Hood hashing} is a technique that was
originally introduced in 1985~\cite{RobinHoodHashing}. The idea behind
Robin Hood hashing is to move already stored elements during
insertions in a way that minimizes the longest possible search
distance.  Robin Hood hashing has regained some popularity in recent
years -- mainly for its interesting theoretical properties and the
possibility to reduce the inherent variance of linear probing.
\emph{Hopscotch hashing}~\cite{HopscotchHashing} is a technique, that
originated in the research of concurrent hash tables, but it can also
manage densely filled hash tables.  Hopscotch hashing also moves
elements within the table to reduce the maximum search distance.
Additionally it introduces an acceleration data structure that is
designed to improve lookup performance especially in cases where the
searched key is not found.

All these publications show that there is a clear interest in
developing hash tables that can be more and more densely filled.
Dynamic hash tables on the other hand seem to be considered a solved
problem.  One paper that takes on the problem of dynamic hash tables
was written by Dietzfelbinger at
al.~\cite{DynamicPerfectHashingUpperAndLowerBounds}.  It actually
predates cuckoo hashing, and much of the attention for space efficient
hashing.  All memory bounds presented are given without tight constant
factors.  The lack of implementations and theory about dense dynamic
hash tables is where we pick up and offer a fast hash table
implementation, that supports dynamic growing with tight space bounds.

\section{Preliminaries}
A hash table is a data-structure for storing key-value-pairs
($\langle key, data \rangle$) and offers the following functionality:
\verb~insert~ -- stores a given key-value pair; \verb~find~ -- given a
key returns if said key was stored, and $\bot$ otherwise; and
\verb~erase~ -- deletes a previously inserted element (if present).

@@

Throughout this paper $n$ denotes the number of elements and $m$ the
number of cells ($m > n$) in any given hash table.  We define the load
factor as $load = n/m$ and the fraction of empty cells $\varepsilon =
1-load$.  Tables can usually only work efficiently up to a certain
load factor.  Above that, operations get slower and/or have a
possibility to fail.  When implementing a hash table one has to decide
between storing elements directly in the hash table -- \emph{Closed
  Hashing} -- or storing pointers to elements -- \emph{Open
  Hashing}. This has an immediate impact on the necessary memory:
\begin{align*}
  mem_{closed}(n,m) &\geq m\cdot size(element)\\
  mem_{open}  (n,m) &\geq m\cdot size(pointer) + n\cdot size(element)\\
\end{align*}

For large elements ($\gg size(pointer)$), one can use a non-space
efficient hash table with open hashing to reduce the relevant memory
factor.  Therefore, we restrict ourselves to the common and more interesting case of
elements whose size is comparable to that of a pointer.  For our
experiments we use 128bit elements (64bit keys and 64bit values).  In
this case, open hashing introduces a significant memory
overhead (at least $\times1.5$). For this reason, we only consider closed
hash tables. For them, the
memory efficiency is directly dependent on their load.  To
reach high fill degrees with closed hashing tables, we have to employ
\emph{open addressing} techniques which means that elements are not
stored in predetermined cells, but can be stored in one of several
options (e.g. linear probing, or cuckoo hashing).

\subsection{$\alpha$-Space Efficient Hash Tables}
\subparagraph*{Static}
\label{sec:pre_staticspace}
We call a hashing technique $\alpha$-space efficient when it can work
effectively using at most $\alpha \cdot n_{\textit{curr}} \cdot \textit{size}(\textit{element}) + O(1)$ memory. In
this case we define working efficiently as having average insertion
times in $O(\frac{1}{\varepsilon})$. This is a natural bound
since it is the expected number of fully random probes needed to hit
an empty cell.

@@

In many closed hashing techniques (e.g.~linear probing, cuckoo
hashing) cells are the same size as elements. Therefore, being
$\alpha$-space efficient is the same as operating with a load factor
of $\alpha^{-1} = n/m$.
%(because $\alpha\cdot n \cdot size(element) = m\cdot size(element)$).
Because of this, we will mostly talk about the load factor of a table
instead of its memory usage.

@@

\paragraph*{Dynamic}
The definition of a space efficient hashing technique given above is
specifically targeted for statically sized hash tables.  We call an
implementation dynamically $\alpha$-space efficient if an instantiated
table can grow arbitrarily large over its original capacity, while
remaining smaller than $\alpha\cdot n_{\max}\cdot
\textit{size}(\textit{element}) + O(1)$ at all times.
%\[mem_{curr} \leq \alpha\cdot max(n_{curr})\cdot size(element)\]

One problem for many implementations of space efficient hash tables is
the migration.  During a normal full table migration, both the
original table, and the new table are allocated.  This takes
$\textit{m}_{\textit{new}} + \textit{m}_{\textit{old}}$
cells. Therefore, a normal full table migration is never better than 2-space
efficient.  The only option to perform a full table migration with
less memory is to increase the memory in place (see
Section~\ref{sec:dys_inplace} and Section~\ref{sec:exp_competitor}).

Similar to static $\alpha$-space efficiency, we will mostly talk about
the \emph{minimum load factor} instead of $\alpha$.  We note that this
definition of dynamic $\alpha$-space efficient hash tables does not
enforce size reduction due to the deletion of elements. However it
enforces that cells previously used by deleted elements must be
\emph{reused}.

\subsection{Cuckoo Hashing}
\label{sec:pre_cuckoo}
Cuckoo hashing is a technique to resolve hash conflicts in a closed
hashing hash table. Its main draw is that it guarantees constant
lookup times even in densely filled tables.  The distinguishing
technique of cuckoo hashing is that $H$ hash functions ($h_1, ... ,
h_H$) are used to compute $H$ independent positions. Each element is
stored in one of its positions.  Even if all positions are occupied
one can often move elements to create space for the current
element. We call this process \emph{displacing} elements.

TODO MAYBE ALREADY GRAPH VIEW

Bucket cuckoo hashing is a variant, where the cells of the hash table
are grouped into buckets of size $B$ ($m/B$ buckets).  Each element
assigned to one bucket can be stored in any of the bucket's cells .
Using buckets one can drastically increase the number of elements that
can be displaced to make room for a new one, thus decreasing the
expected length of displacement paths.

\emph{Find} and \emph{remove} operations have a guaranteed constant
running time. Independent from the tables density there are $H$
buckets -- $H\cdot B$ cells -- that have to be searched to find an
element.

During an \emph{insert} the element is hashed to $H$ buckets.  If at least
one of those buckets has an empty cell, we store the element in the
bucket with the most free space.  When all buckets are full we
have to move elements within the table such that a free cell becomes
available.

\label{sec:cuckoo_graph}
To visualize the problem of displacing elements, one can think of the
directed graph implicitly defined by the hash table.  Each bucket
corresponds to a node and each element induces an edge between
the bucket it is stored in and its $H-1$ alternate buckets.  To insert
an element into the hash table we have to find a path from one of its
associated buckets, to a bucket, that has space remaining.  Then we
move elements along this path to make room in the initial bucket.
The two common techniques to find such paths are \emph{random walks} and
\emph{breadth first searches}.

\section{DySECT (Dynamic Space Efficient Cuckoo Table)}
A commonly used growing technique is to double the size of a hash
table by migrating all its elements into a table with twice its
capacity.  This is of course not memory efficient.  The idea behind
our hash table architecture is to double only parts of the overall
data-structure.  This increases the space in part of our
data-structure without changing the rest.  We then use cuckoo
displacement techniques to make this additional memory reachable from
other parts of the hash table.

\subsection{Overview}
Our DySECT hash table consists of $T$ subtables (shown in TODO
FIGREFRENCE), that in turn consist of buckets that store elements.
Each element has $H$ associated
buckets similar to cuckoo hashing which can be in the same or
in different subtables.  To find a bucket associated with an element $e$,
we compute $e$'s hash value using the appropriate hash function
$h_i(e)$. The hash is then used to compute the subtable and the bucket
within that subtable.

To make this efficient we use powers of two for the number of
subtables ($T = 2^t$), as well as for the individual subtable sizes ($s =
2^x*B$).  Since the number of subtables is constant we can use the
first $t$ bits from the hashed key, to find the appropriate subtable.
From the remaining bits we compute the bucket within the subtable using
a bitmask ($h_i(e)\,\texttt{\&}\,(2^x-1) = h_i(e) \mod 2^x$).

@@

\subsection{Growing}
We grow one subtable by migrating it into a table twice its size
as soon as the (overall) table contains enough elements such that the memory constraint
can be kept during the migration.  We migrate subtables in order
from the first to the last. This ensures that no subtable can be more than
twice as large as any other.

Assume that all subtables have the same size ($s=m/T$). When $\alpha\cdot n >
m+2s$ we can grow the first subtable while keeping the size
constraint.  Doubling the size of a subtable increases the global
number of cells from $m_{old} = T\cdot s$ to $m_{new} = m_{old}+s =
(T+1)\cdot s$ (grow factor $\frac{T+1}{T}$).  Note that all subsequent
growing operations migrate one of the smaller tables, until all tables have the
same size again.  Therefore, each grow until then increases the
overall capacity by the same absolute amount (smaller overall factor TODO EXPLAIN FURTHER?).

The cost of growing a subtable is amortized by all insertions that
happened since the last subtable migration.  There are at least
$\alpha^{-1} \cdot s = \Omega(s)$ insertions between two migrations.
One migration takes $\Theta(s)$ time.  Apart from being amortized, the
migration is cache efficient since it accesses cells in a linear
fashion.  Even in the target table cells are accessed linearly.  We
assign elements to buckets by using bits from their hash value. In the
grown table we use exactly one more bit than before (double the number
of buckets).  This ensures, that all elements from one original
bucket, are split between two buckets in the target table.
Therefore, no bucket can overflow and no displacements are necessary.

In the implicit graph model of the cuckoo table
(Section~\ref{sec:cuckoo_graph}), growing a subtable is equivalent to
splitting each node that represents a bucket within that table. Since
the edges (elements) in the implicitly defined subgraph are not
doubled, the resulting subgraph becomes more sparse, making it easier
to insert elements.

\subsection{Shrinking}
In many use cases automatic shrinking is not necessary.  It worsens
performance by taking time for the migration and making the remaining
table more dense.  Therefore, we forgo automatic shrink operations
whenever enough elements are removed from the table.

If shrinking is necessary it can work similarly to growing. We
replace a subtable with a smaller one by migrating elements from one
to the other.  During this migration we join elements from two buckets
into one. Therefore, it is possible for a bucket to overfill.  We
reinsert these elements at the end of the migration.  Obviously, this
can only apply to at most half the migrated elements.

When automatically triggering the size reduction, one has to make sure
that the migration cost is amortized. Therefore, a grow operation cannot
immediately follow a shrink operation.  When shrinking is enabled we
propose to shrink one subtable when $\alpha\cdot n < m-s'$ elements
($s'$ size of a large table, $m_{new} = m_{old} -s'/2$).
Alternatively, one could implement a \emph{shrink to size} operation that is
explicitly called by the user.

@@

\subsection{Difficulties for the Analysis of our table structure}
There are two factors that impact the performance of our dynamic
table compared to other cuckoo table variants and to other hashing
solutions in general: \textbf{inhomogeneous table resolution} and
\textbf{element imbalance}.
%, and \textbf{population density}.
Both factors influence the maximum load density and the
running times in different ways.

\subparagraph*{Imbalance through Inhomogeneous Table Resolution.}
\label{sec:inhom_res}
By growing subtables individually we introduce a size imbalance
between subtables. Large subtables contain more buckets but the number
of elements hashed to a large subtables is not generally higher than
the number of elements that are hashed to a small subtable.  This
makes it difficult to spread elements evenly among buckets.
Imbalanced bucket fill ratios can lead to longer insertion times.

Assume there are $n$ elements in a hash table with $T$ subtables,
$j$ of which have size $2s$ the others have size $s$. If elements are
spread equally among buckets then all small tables have around
$n/(T+j)$ elements, and the bigger tables have $2n/(T+j)$ elements.
For each table there are about $Hn/T$ elements associated to one of
its buckets.  This shows that having more hash functions can lead to a
better balance.
